---
title: "A complete analysis in five steps (student tutorial)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete analysis in five steps (student tutorial)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5, fig.align = "center"
)
```

This tutorial takes you from a raw variety-trial dataset to a finished report,
using nothing but the built-in example data. It is written for students and
early-career researchers: every step is one short command, and you can copy the
whole thing and run it top to bottom.

The journey is:

> **explore** → **analyse** → **visualise** → **report**

```{r setup}
library(PbStatR)
```

## The data

PbStatR ships example datasets so you never have to hunt for one. We will use
`"met"`, a multi-environment yield trial. For this first tutorial we focus on a
**single environment** (a classic RCBD) to keep things simple.

```{r data}
trial <- pb_data("met")
str(trial)

# keep one environment -> a randomised complete block design (RCBD)
rcbd <- droplevels(subset(trial, env == "E1"))
head(rcbd)
```

We have `gen` (genotype/variety), `rep` (block), and two traits, `yield` and
`height`. That is all you need for a full genetic analysis.

## Step 1 — Explore: which function do I need?

Not sure where to start? `pb_explore()` maps a research goal to the right
function and shows a ready example. Give it a keyword:

```{r explore}
pb_explore("heritability")
```

Call it with no argument to see every goal the package covers, or try other
keywords like `"stability"`, `"correlation"`, `"gwas"`, or `"selection"`.

```{r explore-all, eval = FALSE}
pb_explore()              # the full menu
pb_explain("heritability") # plain-language definition of a term
```

If you prefer a ready-made recipe for your whole study, `pb_workflow()` prints
copy-paste code:

```{r workflow}
pb_workflow("variety_trial")
```

## Step 2 — Analyse: one call does the standard analysis

`pb_analyze()` runs the ANOVA, checks the assumptions, estimates heritability
and the genetic parameters, groups the means with a post-hoc test, and explains
the result in plain language. Everything is also returned for later use.

```{r analyze}
result <- pb_analyze(rcbd, trait = "yield",
                     genotype = "gen", block = "rep")
```

The pieces are all there to inspect:

```{r analyze-parts}
result$genetic_params      # GCV, PCV, heritability, GAM ...
head(result$means)         # ranked means with grouping letters
```

And it produced ready-to-use plots:

```{r analyze-plots}
result$plots$boxplot
```

```{r analyze-plots2}
result$plots$means_bar
```

## Step 3 — Visualise: one figure that summarises the genetics

`pb_genetic_dashboard()` packs the whole genetic summary into a single
attractive figure: genotype means with error bars, a variance-partition donut
with heritability in the centre, the genetic-parameter bars, and a heritability
gauge.

```{r dashboard, fig.height = 6}
pb_genetic_dashboard(rcbd, "yield", "gen", "rep")
```

One picture, the complete story: how the varieties rank, how much of the
variation is genetic, and whether selection is likely to work.

## Step 4 — Dig deeper (optional)

The same data supports much more. A few one-liners:

```{r deeper, eval = FALSE}
# trait relationships
pb_correlation(rcbd, c("yield", "height"))
pb_heatmap(rcbd, c("yield", "height"))

# a publication-ready ANOVA table (any design)
pb_aov_design(rcbd, "yield", design = "rcbd",
              treatment = "gen", block = "rep")

# post-hoc mean comparison on its own
pb_posthoc(rcbd, "yield", "gen", "rep", method = "tukey")$groups
```

And if you use the **whole** multi-environment dataset, you unlock stability
analysis:

```{r met, eval = FALSE}
pb_gxe(trial, gen = "gen", env = "env", rep = "rep", trait = "yield")
ranks <- pb_stability_ranks(trial, "gen", "env", "rep", "yield",
                            include_waasb = FALSE)
pb_plot_stability(ranks)
```

## Step 5 — Report: one document to share

When the analysis is done, `pb_report()` renders everything — summary, ANOVA,
assumptions, genetic parameters, mean grouping, and (for multi-environment
data) the stability section — into a single HTML or Word file.

```{r report, eval = FALSE}
# HTML report for the single environment
pb_report(rcbd, "yield", genotype = "gen", block = "rep",
          format = "html")

# a fuller Word report using all environments (adds the G x E section)
pb_report(trial, "yield", genotype = "gen", block = "rep", env = "env",
          format = "word", title = "Yield trial 2024")
```

That is the entire workflow: **explore → analyse → visualise → report**, four
commands from raw data to a shareable document.

## Where to go next

- `pb_help()` — the categorised menu of everything in the package.
- `vignette("PbStatR-beginners")` — a gentler first look.
- `vignette("PbStatR-gwas-stability")` — worked GWAS and stability examples.
- `pb_explore("<your goal>")` — whenever you are unsure what to run next.
