---
title: "Worked examples: GWAS and combined stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked examples: GWAS and combined stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup}
library(PbStatR)
```

This vignette runs two complete workflows on the datasets bundled with the
package, so you can reproduce them exactly.

## Part 1 — GWAS with the example marker data

The package ships a small GWAS-ready dataset: 200 individuals, 500 SNPs on 10
chromosomes, with eight SNPs carrying real additive effects.

```{r}
geno  <- pb_data("geno")    # Taxa + 500 SNP columns (0/1/2)
map   <- pb_data("map")     # SNP, Chr, Pos
pheno <- pb_data("pheno")   # Taxa, trait
dim(geno); head(map)
```

### Tidy the marker map and plot positions

```{r}
tidy_map <- pb_marker_map(map)   # sorts by Chr/Pos, adds cumulative positions
head(tidy_map)
```

### A quick association scan without external packages

Even without a dedicated GWAS engine installed, you can run a simple
marker-by-marker regression to see the workflow. (For production analyses use
`pb_gwas_rmvp()` or `pb_gwas_gapit()`.)

```{r}
M <- as.matrix(geno[, -1])       # drop Taxa column
y <- pheno$trait
pvals <- apply(M, 2, function(snp) {
  if (stats::sd(snp) == 0) return(NA_real_)
  summary(stats::lm(y ~ snp))$coefficients[2, 4]
})
```

Now the Manhattan and QQ plots come straight from the package:

```{r}
pb_manhattan(tidy_map, pvals, threshold = 0.05 / length(pvals))
pb_qqplot(pvals)
```

Did we recover the known causal SNPs? Compare the top hits to the truth:

```{r}
top_hits <- names(sort(pvals))[1:10]
truth <- pb_data_causal()
intersect(top_hits, truth)     # causal SNPs among the top 10
```

### Running a full engine (if installed)

```{r, eval = FALSE}
# rMVP (GLM / MLM / FarmCPU in one call)
res <- pb_gwas_rmvp(phe = pheno, geno = t(M), map = tidy_map[, 1:3],
                    methods = c("MLM", "FarmCPU"))

# GAPIT
res2 <- pb_gwas_gapit(Y = pheno, GD = geno, GM = map, model = "MLM")
```

## Part 2 — Genotype x environment: every method, one table

The `pbs_met` trial has 20 genotypes across 5 environments.

```{r}
met <- pb_data("met")
str(met)
```

Estimate GxE in all the standard ways at once:

```{r}
gxe <- pb_gxe(met, gen = "gen", env = "env", rep = "rep", trait = "yield",
              methods = c("anova", "regression", "shukla", "ecovalence"))
gxe$joint_anova$gxe_pvalue     # is GxE significant?
head(gxe$regression)           # Finlay-Wilkinson bi and s2di
```

### One combined ranking

`pb_stability_ranks()` pulls the statistics together, ranks each, and orders
genotypes by an overall stability-and-performance score (rank 1 = best):

```{r}
ranks <- pb_stability_ranks(met, "gen", "env", "rep", "yield",
                            include_waasb = FALSE)
head(ranks[, c("Genotype", "Mean", "bi", "s2di",
               "Shukla_sigma2", "Ecovalence", "overall_rank")])
```

```{r}
pb_plot_stability(ranks)
```

The ideal genotypes sit high in mean performance with a low stability rank —
the lower-right of the plot. With `metan` installed, set `include_waasb = TRUE`
to add the WAASB index to the same table.
