---
title: "PbStatR for beginners"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PbStatR for beginners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

This guide is written for students who are new to R and to plant breeding
statistics. If you can load a data frame, you can use PbStatR.

```{r setup}
library(PbStatR)
```

## See what the package can do

```{r, eval = FALSE}
pb_help()
```

## The one-call analysis: `pb_analyze()`

Suppose you evaluated 6 varieties in 3 blocks (an RCBD) and measured yield.
`pb_analyze()` does the ANOVA, checks assumptions, estimates heritability and
runs a Tukey grouping — then explains the result in plain language.

```{r}
set.seed(1)
trial <- data.frame(
  variety = rep(paste0("V", 1:6), each = 3),
  block   = rep(1:3, times = 6),
  yield   = c(40, 41, 39, 45, 46, 44, 50, 49, 51,
              42, 43, 41, 38, 37, 39, 47, 48, 46)
)

result <- pb_analyze(trial, trait = "yield",
                     genotype = "variety", block = "block")
```

Everything is returned for later use:

```{r, eval = FALSE}
result$genetic_params   # GCV, PCV, heritability, GAM ...
result$means            # ranked means with grouping letters
result$plots$boxplot    # a ggplot you can print or save
result$plots$means_bar
result$plots$diagnostics
```

## A quick descriptive look

```{r}
pb_summary(trial, traits = "yield", genotype = "variety")
```

## Designing the field first

Before you collect data, lay out the experiment. With FielDHub you also get a
plot map to take to the field:

```{r, eval = FALSE}
layout <- field_rcbd(t = 6, reps = 3)
head(layout$field_book)
layout$field_map        # ggplot of the field
```

## Where to go next

- Post-hoc tests: `pb_posthoc()`, assumptions: `pb_assumptions()`
- Multivariate: `pb_pca()`, `pb_hcpc()`, heatmaps: `pb_heatmap()`
- Stability: `pb_ammi()`, `pb_waasb()`
- Genomic selection: `pb_blup()`, `pb_gblup()`
- Simulation: `pb_sim_founders()`, `pb_sim_pipeline()`

Each has its own help page — type `?pb_analyze` to try it.
